### 2023-07-20 chapters/paper_modelagnostic.tex
Revise skipping

\chapter[model-agnostic out-of-distribution detection using combined statistical tests]{Model-Agnostic Out-of-Distribution Detection Using Combined Statistical Tests}
\label{chp:paper-modelagnostic}
\ifthenelse{\equal{\skippaperbody}{true}}{}{

@@

The ability to recognise when data are anomalous, i.e.\@ if they originate from a distribution different from that of the training data, is a necessary property for machine learning models for safe and reliable applications in the real world. Historically, \citet{bishop1994novelty} proposed to use a one-sided threshold on the log-likelihoods of a learned model as a decision rule to identify outliers in a dataset. However, recently, \citet{nalisnick2018deep, hendrycks2018deep} showed that state-of-the-art deep generative models (DGMs) failed in this task, assigning higher a likelihood to out-of-distribution (OOD) data than in-distribution data. Most of the recent works focused on proposing new test statistics to alleviate the problem of using the plain likelihood, see \cref{sec:related_works} for details.

We believe that OOD detection should be formulated as statistical hypothesis testing \citep{nalisnick2019detecting, ijcai2021-292, haroush2021statistical}. Since the power of a single test depends on the out-distribution \citep{zhang2021understanding}, we propose to approach this problem by using a combination of multiple statistical tests. %While this does not solve completely the problem, because the power of the combined tests will also depend on the out-distribution, we think that we can get better performance than using a single-statistics especially in situations where one statistic fails. 
While the power of the combined test also depends on the out-distribution, we hypothesise that the combined test empirically will perform better, especially in situations where one of the statistics fails.
%
%Furthermore, the use of the statistical testing framework, allows us to correct for the multiple comparisons problem when identifying outliers in a dataset by controlling the number of type I errors through the false discovery rate (FDR).
Furthermore, the use of the statistical testing framework has several advantages. Since we obtain a $p$-value, it is more natural deciding on a threshold as this corresponds to the significance level. In addition to that, it also allow us to correct for the multiple comparisons problem when identifying outliers in a dataset by controlling the number of Type I errors through the false discovery rate (FDR).

In summary, our contributions are the following:
\begin{itemize}
    \item We illustrate the benefits of combining multiple statistical tests to perform OOD detection with DGMs using well-established methods. This allows for a proper decision procedure to control the FDR in a real outlier detection setting.
    \item We revisit some proposed detection scores and highlight their alternative formulation as classical significance tests.
    \item Empirically we show the complementarity of the typicality and the score statistics and that their combination leads to a robust score for anomaly detection. %We evaluate our proposed method with two PixelCNN++ \cite{salimans2017pixelcnn++}, two Glow models \cite{kingma2018glow} and one Hierarchical-VAE (HVAE)\cite{kingma2013auto}.
\end{itemize}
%

@@

Note that in this problem formulation, the case $n=1$ corresponds to the situation where we need to decide if a \emph{single} data point is out-of-distribution. This hardest setting will be of particular interest, and this is also the main focus of recent work, see \cref{sec:related_works}.

\subsection{Parametric tests for out-of-distribution detection}\label{sec:parametric}
The typical approach is to consider a parametric family $(p_\theta)_{\theta \in  \Theta}$ of probability densities over $\mathcal{X}$ and learn a suitable $\theta_0 \in  \Theta$ using any inference technique, for example maximum likelihood, and the clean data $x_1,\ldots,x_m$. Depending on the input domain, $(p_\theta)_{\theta \in  \Theta}$ could be composed of DGMs (in that case, $\theta$ would be neural network weights) or Gaussian mixture models (in that case, $\theta$ would be composed of means, covariances, and proportions). The question we wish to answer may then be phrased: \emph{is $p_{\theta_0}$ an appropriate model for $\tilde{x}_1,\ldots,\tilde{x}_n$?}
%We consider some data of interest that live in a space $\mathcal{X}$. We are given a parametric family $(p_\theta)_{\theta \in  \Theta}$ of probability densities over $\mathcal{X}$. For instance $(p_\theta)_{\theta \in  \Theta}$ could be composed of DGMs (in that case, $\theta$ would be neural network weights) or Gaussian mixture models (in that case, $\theta$ would be composed of means, covariances, and proportions).
%
%Upon analysis of an initial, in-distribution training dataset, we learn a suitable $\theta_0 \in  \Theta $ using any inference technique, for example maximum likelihood. We are now given additional data points $\tilde{x}_1,\ldots,\tilde{x}_n$ and wonder whether or not they are likely to come from the same distribution that generated our training dataset. The case $n=1$ corresponds to the case where we need to decide if a \emph{single} data point is out-of-distribution; this hardest setting will be of particular interest.
%The question we wish to answer may then be phrased: \emph{is $p_{\theta_0}$ an appropriate model for $\tilde{x}_1,\ldots,\tilde{x}_n$?}

@@

Many tests have been proposed for this purpose. The three most famous are the \emph{likelihood ratio test} of \citet{neyman1928use}, Rao's \citeyearpar{rao1948large} \emph{score test},
%(also called \emph{Lagrange multiplier test}, mostly by econometricians),
and \emph{the Wald test} \citep{wald1943tests}. These three classics are nicely reviewed by \citet{buse1982likelihood} or by \citet{rao2005score}, who called them the ``Holy Trinity''. A recent and interesting one is the \emph{gradient test} of \citet{terrell2002gradient}, which is reviewed in great detail in Lemonte's \citeyearpar{lemonte2016gradient} monograph.

Let us review the statistics of these four tests:
\begin{itemize}
    \item likelihood ratio statistic is $S_{LR} = 2 ( \ell ( \hat{\theta}) -\ell ( \theta_0))$,
    \item Wald statistic is $S_{W} = (\hat{\theta} - \theta_0)^T I(\hat{\theta}) (\hat{\theta} - \theta_0)$,
    \item score statistic is $S_{S} = \nabla \ell ( \theta_0)^T I(\theta_0)^{-1}  \nabla \ell ( \theta_0)$,
    \item gradient statistic is $S_{G} = \nabla \ell ( \theta_0)^T (\hat{\theta} - \theta_0)$,
\end{itemize}
where $ \ell ( \theta) = \log p_\theta (\tilde{x}_1,\ldots,\tilde{x}_n)$ is the likelihood function, 
$I({\theta}) = \mathbb{E}_{p_\theta}[\nabla \ell ( \theta)\nabla \ell ( \theta)^T]$ is the Fisher information matrix (FIM), and $ \hat{\theta} \in \argmax_{\theta \in \Theta} \ell ( \theta)$.

The likelihood ratio statistic, the Wald statistic and the gradient statistic all require to fit a model on the additional datapoints $\tilde{x}_1,\ldots,\tilde{x}_n$ in order to compute either $\ell(\hat{\theta})$ or $\hat{\theta}$. 
%
%In our setting, if we want to use one of those statistics as an OOD score for a single example, we should fit a DGM on that single datapoint. \cite{xiao2020likelihood} try to do that in the context of VAEs by re-fitting only the inference network, or encoder, of a VAE to every single additional example. This is the typical thing to do when dealing with out-of-sample data in VAEs, as argued by \cite{mattei2018refit, cremer2018inference}. 
In our setting, if we want to use one of those statistics as an OOD score for a single example, we should fit a DGM on that single datapoint. \cite{xiao2020likelihood} did this for a variational autoencoder (VAE, \citealp{kingma2013auto,rezende2014stochastic}) by only re-fitting inference network (or encoder) to the additional example, which is a typical approach to dealing with out-of-sample data in VAEs, as argued by \cite{cremer2018inference} and \citet{mattei2018refit}.
However, much of the recent works in the literature \citep{ren2019likelihood, schirrmeister2020understanding, serra2019input} mainly focus on deriving different versions of what they call a likelihood ratio statistic.

%In our setting, if we want to use one of those statistics as an OOD score for a single example, we should fit a DGM on that single datapoint, which barely sounds serious. Most of the recent works in the literature mainly focus on deriving different versions of what they call a likelihood ratio statistic. Among those that trained a likelihood-based model on $\tilde{x}_1,\ldots,\tilde{x}_n$, we note that the work by \cite{xiao2020likelihood} is the one that is closer to the real definition of the likelihood ratio statistic given by \cite{neyman1928use}. Indeed, they propose to compute $\ell(\hat{\theta})$ by refitting the inference network of a VAE to every single additional example as proposed in \cite{mattei2018refit, cremer2018inference}. 
%While a similar approach could potentially be used to compute also the gradient statistic and the Wald statistic, this will be possible only for a VAE model. 
%
We tried to derive a general way to compute both the Wald statistic and the gradient statistic, by computing $ \hat{\theta}$ with a few steps of a gradient-based optimization algorithm initialized at $ \theta_0$, but this resulted in a very unstable update leading to computational issues (results not shown). Therefore, in this work we focus on studying the relevance of the score statistic for performing out-of-distribution detection since it is the only statistic that does not require fitting an additional model to the OOD data.

\subsection{Maximum mean discrepancy for out-of-distribution detection}\label{sec:MMD}
Another way of approaching out-of-distribution detection from a testing perspective is through a \emph{two-sample test}. Denoting $p_\textup{data}$ the true training data distribution, the goal is to test $\mathcal{H}_0: \tilde{x}_1,\ldots,\tilde{x}_n \sim p_\textup{data}$ against $\mathcal{H}: \tilde{x}_1,\ldots,\tilde{x}_n \not\sim p_\textup{data}$, where the alternative hypothesis $\mathcal{H}$ again is that the test points are OOD.

A popular way of building statistics for two-sample tests is to use a measure of distance between $p_\textup{data}$ and the distribution of $\tilde{x}_1,\ldots,\tilde{x}_n$. The key idea here will be to use the trained generative model to build this measure of distance. To this end, we will use the \emph{maximum mean discrepancy (MMD)} of \citet{gretton2012kernel}, which is a kernel-based measure of distance. Then, $p_\theta$ will be used to specify an appropriate kernel.

More specifically, given a kernel whose feature map is $\Phi : \mathcal{X} \to \mathcal{H}$, the MMD between two distributions $P$ and $Q$ over $\mathcal{X}$ is defined as
\begin{equation}
    \textup{MMD}_\Phi(P,Q) = \| E_{X\sim P}[\Phi(X)] - E_{Y\sim Q}[\Phi(Y)] \|_\mathcal{H}. \hspace{-.1em}
\end{equation}
In our context, the test statistics will be of the form
\begin{multline}
    \label{eq:mmd}
    \textup{MMD}_\Phi\left(\frac{1}{m}\sum_{i=1}^{m} x_i,\frac{1}{n}\sum_{i=1}^n \tilde{x}_i\right) = \\\left\|\frac{1}{m}\sum_{i=1}^{m} \Phi(x_i) -  \frac{1}{n}\sum_{i=1}^{n} \Phi(\tilde{x}_i)\right\|_\mathcal{H},
\end{multline}
where $\Phi$ is a kernel feature map built using the generative model and $x_1,\ldots,x_m$ is the training data, i.e.\@ samples from  $p_\textup{data}$. When $\mathcal{H}$ is a simple finite-dimensional Hilbert space and $\Phi$ can be computed easily, then \eqref{eq:mmd} can be computed by going through the data and computing the means in an online fashion.% While the definition seems to imply that we can only perform group-sample OOD detection, it is straightforward to notice that any test statistic of the form of \cref{eq:mmd} can be used also for single-test OOD detection when $n=1$.%by just substituting  $\frac{1}{n}\sum_{i=1}^{n} \Phi(\tilde{x}_i)$ with simply $\Phi(\tilde{x})$.

@@

\paragraph{The Fisher kernel} An important example of kernel based on a generative model is the \emph{Fisher kernel} of \citet{jaakkola1999exploiting}. The embedding of this kernel is the Fisher score 
\begin{equation}
    \label{eq:Fisher_score}
    \Phi_\textup{Fisher} (x) = I(\theta)^{-\frac{1}{2}} \nabla \log p_\theta (x),
\end{equation}

and the corresponding reproducing kernel Hilbert space norm is just the $\ell_2$ norm: $|| \cdot || _\mathcal{H} = || \cdot ||_2$. In the case of the Fisher kernel, this means that \cref{eq:mmd} becomes:
\begin{multline}
    \label{eq:mmd_fisher}
    \textup{MMD}_{\Phi_\textup{Fisher}}\left(\frac{1}{m} \sum_{i=1}^{m} x_i,\frac{1}{n}\sum_{i=1}^n \tilde{x}_i\right) = \\ \left\|\frac{I(\theta)^{-\frac{1}{2}}}{m} \sum_{i=1}^{m}  \nabla \log p_\theta (x_i) - \frac{I(\theta)^{-\frac{1}{2}}}{n}  \sum_{i=1}^{n}  \nabla \log p_\theta (\tilde{x}_i)\right\|_2.
\end{multline}
We will see later that MMD with a Fisher kernel is closely related to the score statistic. In \cref{appendix:mahalanobis}, we additionally show that another popular OOD metric known as the \emph{Mahalanobis score} \citep{lee2018simple} can be interpreted as a MMD statistic with a certain Fisher kernel.

\paragraph{The typicality kernel} A very simple approach of embedding the data using $p_\theta$ is to choose $\Phi_\textup{Typical}(x) = \log p_\theta (x)$. Then, MMD is exactly equivalent to the \emph{typicality test statistic} of \citet{nalisnick2019detecting}, although this connection was not explicitly stated  by \citet{nalisnick2019detecting}. Because of this, we call the kernel $k(x,y) = \log p_\theta (x) \cdot \log p_\theta (y)$ the \emph{typicality kernel}.  %We could also call this "new" kernel the "likelihood kernel".
%While $\Phi_\textup{Typical}$ is not as well motivated as the Fisher kernel, we found that it generally gives good results. 
While $\Phi_\textup{Typical}$ is not as well motivated as a kernel as $\Phi_\textup{Fisher}$, the concepts of typicality and typical set can be used to explain unintuitive behaviours of probability distributions in high-dimensional space as highlighted by \cite{nalisnick2018deep}. We also found that using this kernel generally gives good results for OOD tasks. An interesting analysis that we d not consider in this paper would be to study the properties of this kernel.

In general, neither of these two kernels are characteristic, meaning that our MMD can be zero even if the distributions are not identical. This could be solved by combining them with a characteristic kernel, as in \cite{liu2020learning}, at the price of including a new hyperparameter.

@@

%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%  SECTION 3   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Combining different test statistics}
\label{sec:combination_p_values}
For single-sample OOD detection, \citet{zhang2021understanding} proved that there is not a single statistic that is constantly better compared to all the possible alternatives of interest. For this reason, we believe that using a combination of different test statistics should lead to an overall better OOD detection in settings where a single statistic might fail. Assume we compute $k$ different test statistics $T_1,\dots, T_k$, each testing $\mathcal{H}_0$ against $\mathcal{H}$ as defined in \cref{sec:parametric,sec:MMD}. The goal is to combine these different tests into a single statistical test that ideally will perform better than the initial single tests. However, different tests can have different magnitudes and they can differ also in the direction of out-of-distribution detection, i.e.\@ for some statistics having a higher values is associated with being OOD, while for other smaller values are OOD. This makes a combination non-trivial.

\citet{morningstar2021density} proposed the density of states estimator (DoSE) to overcome this problem. They only focused on the single-sample detection task, i.e.\@ $n=1$ following our problem formulation. Their idea is to fit different nonparametric density estimators, such as a kernel-density estimator (KDE) or a one-class support vector machines (SVM), for each different statistic $T_1,\dots, T_k$ by using the values computed on the training set examples. For a single test example, $\tilde{x}_1$, they first compute $T_1,\dots, T_k$ and then combine those statistics by summing the different KDEs log-density. %This approach can be used when combining both one-sided and two-sided test statistic. 
While this approach can be used for any type of statistic, and thus is more general, it uses less prior information. Indeed, if we use only statistics that are truly one-sided, then we assume that a method that leverages the true nature of the statistics should work better. In addition to that, fitting a KDE introduces an additional hyperparameter.

In our work, instead, we propose a different approach and leverage the fact that we use only one-sided test statistics. This setting is a well-studied problem in the literature both for independent \citep{fisher1925statistical, folks1971asymptotic} and dependent one-sided test statistics \citep{brown1975400, wilson2019harmonic}. All these approaches rely on the computation of $p$-values of each statistic for the test set $\tilde{x}_1,\ldots,\tilde{x}_n$. This corresponds to computing
$p_j = \textup{Pr}(T_j > t_j \mid \mathcal{H}_0)$, i.e.\@ the probability that the $j$'th test is bigger than the observed value under the null hypothesis $\mathcal{H}_0$, where we assume that each $T_j$ has a continuous distribution. Using $p$-values also solves the problem of the statistics having different scales. Indeed, $p$-values transform the different test statistics into the unit interval.

>>>>

\chapter[model-agnostic out-of-distribution detection using combined statistical tests]{Model-Agnostic Out-of-Distribution Detection Using Combined Statistical Tests}
\label{chp:paper-modelagnostic}
\ifthenelse{\equal{\skippapers}{true}}{}{

@@

The ability to recognise when data are anomalous, i.e. if they originate from a distribution different from that of the training data, is a necessary property for machine learning models for safe and reliable applications in the real world. Historically, \textcite{bishop_novelty_1994} proposed to use a one-sided threshold on the log-likelihoods of a learned model as a decision rule to identify outliers in a dataset. However, recently, \textcite{nalisnick_deep_2019,hendrycks_deep_2019} showed that state-of-the-art deep generative models (DGMs) failed in this task, assigning higher a likelihood to out-of-distribution (OOD) data than in-distribution data. Most of the recent works focused on proposing new test statistics to alleviate the problem of using the plain likelihood, see \cref{sec_modelagnostic:related_works} for details.

We believe that OOD detection should be formulated as statistical hypothesis testing \parencite{nalisnick_detecting_2019, ahmadian_likelihoodfree_2021, haroush_statistical_2021}. Since the power of a single test depends on the out-distribution \parencite{zhang_understanding_2021}, we propose to approach this problem by using a combination of multiple statistical tests. %While this does not solve completely the problem, because the power of the combined tests will also depend on the out-distribution, we think that we can get better performance than using a single-statistics especially in situations where one statistic fails. 
While the power of the combined test also depends on the out-distribution, we hypothesise that the combined test empirically will perform better, especially in situations where one of the statistics fails.
%
%Furthermore, the use of the statistical testing framework, allows us to correct for the multiple comparisons problem when identifying outliers in a dataset by controlling the number of type I errors through the false discovery rate (FDR).
Furthermore, the use of the statistical testing framework has several advantages. Since we obtain a $p$-value, it is more natural deciding on a threshold as this corresponds to the significance level. In addition to that, it also allow us to correct for the multiple comparisons problem when identifying outliers in a dataset by controlling the number of Type I errors through the false discovery rate (FDR).

In summary, our contributions are the following:
\begin{itemize}
    \item We illustrate the benefits of combining multiple statistical tests to perform OOD detection with DGMs using well-established methods. This allows for a proper decision procedure to control the FDR in a real outlier detection setting.
    \item We revisit some proposed detection scores and highlight their alternative formulation as classical significance tests.
    \item Empirically we show the complementarity of the typicality and the score statistics and that their combination leads to a robust score for anomaly detection. %We evaluate our proposed method with two PixelCNN++ \textcite{salimans_pixelcnn_2017}, two Glow models \textcite{kingma_glow_2018} and one Hierarchical-VAE (HVAE)\textcite{kingma_autoencoding_2014}.
\end{itemize}
%

@@

Note that in this problem formulation, the case $n=1$ corresponds to the situation where we need to decide if a \emph{single} data point is out-of-distribution. This hardest setting will be of particular interest, and this is also the main focus of recent work, see \cref{sec_modelagnostic:related_works}.

\subsection{Parametric tests for out-of-distribution detection}\label{sec_modelagnostic:parametric}
The typical approach is to consider a parametric family $(p_\theta)_{\theta \in  \Theta}$ of probability densities over $\mathcal{X}$ and learn a suitable $\theta_0 \in  \Theta$ using any inference technique, for example maximum likelihood, and the clean data $x_1,\ldots,x_m$. Depending on the input domain, $(p_\theta)_{\theta \in  \Theta}$ could be composed of DGMs (in that case, $\theta$ would be neural network weights) or Gaussian mixture models (in that case, $\theta$ would be composed of means, covariances, and proportions). The question we wish to answer may then be phrased: \emph{is $p_{\theta_0}$ an appropriate model for $\tilde{x}_1,\ldots,\tilde{x}_n$?}
%We consider some data of interest that live in a space $\mathcal{X}$. We are given a parametric family $(p_\theta)_{\theta \in  \Theta}$ of probability densities over $\mathcal{X}$. For instance $(p_\theta)_{\theta \in  \Theta}$ could be composed of DGMs (in that case, $\theta$ would be neural network weights) or Gaussian mixture models (in that case, $\theta$ would be composed of means, covariances, and proportions).
%
%Upon analysis of an initial, in-distribution training dataset, we learn a suitable $\theta_0 \in  \Theta $ using any inference technique, for example maximum likelihood. We are now given additional data points $\tilde{x}_1,\ldots,\tilde{x}_n$ and wonder whether or not they are likely to come from the same distribution that generated our training dataset. The case $n=1$ corresponds to the case where we need to decide if a \emph{single} data point is out-of-distribution; this hardest setting will be of particular interest.
%The question we wish to answer may then be phrased: \emph{is $p_{\theta_0}$ an appropriate model for $\tilde{x}_1,\ldots,\tilde{x}_n$?}

@@

Many tests have been proposed for this purpose. The three most famous are the \emph{likelihood ratio test} of \textcite{neyman_use_1928}, Rao's \citeyearpar{rao_large_1948} \emph{score test},
%(also called \emph{Lagrange multiplier test}, mostly by econometricians),
and \emph{the Wald test} \parencite{wald_tests_1943}. These three classics are nicely reviewed by \textcite{buse_likelihood_1982} or by \textcite{rao_score_2005}, who called them the ``Holy Trinity''. A recent and interesting one is the \emph{gradient test} of \textcite{terrell_gradient_2002}, which is reviewed in great detail in Lemonte's \citeyearpar{lemonte_gradient_2016} monograph.

Let us review the statistics of these four tests:
\begin{itemize}
    \item likelihood ratio statistic is $S_{LR} = 2 ( \ell ( \hat{\theta}) -\ell ( \theta_0))$,
    \item Wald statistic is $S_{W} = (\hat{\theta} - \theta_0)^T I(\hat{\theta}) (\hat{\theta} - \theta_0)$,
    \item score statistic is $S_{S} = \nabla \ell ( \theta_0)^T I(\theta_0)^{-1}  \nabla \ell ( \theta_0)$,
    \item gradient statistic is $S_{G} = \nabla \ell ( \theta_0)^T (\hat{\theta} - \theta_0)$,
\end{itemize}
where $ \ell ( \theta) = \log p_\theta (\tilde{x}_1,\ldots,\tilde{x}_n)$ is the likelihood function, 
$I({\theta}) = \mathbb{E}_{p_\theta}[\nabla \ell ( \theta)\nabla \ell ( \theta)^T]$ is the Fisher information matrix (FIM), and $ \hat{\theta} \in \arg\max_{\theta \in \Theta} \ell ( \theta)$.

The likelihood ratio statistic, the Wald statistic and the gradient statistic all require to fit a model on the additional datapoints $\tilde{x}_1,\ldots,\tilde{x}_n$ in order to compute either $\ell(\hat{\theta})$ or $\hat{\theta}$. 
%
%In our setting, if we want to use one of those statistics as an OOD score for a single example, we should fit a DGM on that single datapoint. \textcite{xiao_likelihood_2020} try to do that in the context of VAEs by re-fitting only the inference network, or encoder, of a VAE to every single additional example. This is the typical thing to do when dealing with out-of-sample data in VAEs, as argued by \textcite{mattei2018refit, cremer2018inference}. 
In our setting, if we want to use one of those statistics as an OOD score for a single example, we should fit a DGM on that single datapoint. \textcite{xiao_likelihood_2020} did this for a variational autoencoder (VAE, \citealp{kingma_autoencoding_2014,rezende_stochastic_2014}) by only re-fitting inference network (or encoder) to the additional example, which is a typical approach to dealing with out-of-sample data in VAEs, as argued by \textcite{cremer_inference_2018} and \textcite{mattei_refit_2018}.
However, much of the recent works in the literature \parencite{ren_likelihood_2019, schirrmeister_understanding_2020, serra_input_2020} mainly focus on deriving different versions of what they call a likelihood ratio statistic.

%In our setting, if we want to use one of those statistics as an OOD score for a single example, we should fit a DGM on that single datapoint, which barely sounds serious. Most of the recent works in the literature mainly focus on deriving different versions of what they call a likelihood ratio statistic. Among those that trained a likelihood-based model on $\tilde{x}_1,\ldots,\tilde{x}_n$, we note that the work by \textcite{xiao_likelihood_2020} is the one that is closer to the real definition of the likelihood ratio statistic given by \textcite{neyman_use_1928}. Indeed, they propose to compute $\ell(\hat{\theta})$ by refitting the inference network of a VAE to every single additional example as proposed in \textcite{mattei2018refit, cremer2018inference}. 
%While a similar approach could potentially be used to compute also the gradient statistic and the Wald statistic, this will be possible only for a VAE model. 
%
We tried to derive a general way to compute both the Wald statistic and the gradient statistic, by computing $ \hat{\theta}$ with a few steps of a gradient-based optimization algorithm initialized at $ \theta_0$, but this resulted in a very unstable update leading to computational issues (results not shown). Therefore, in this work we focus on studying the relevance of the score statistic for performing out-of-distribution detection since it is the only statistic that does not require fitting an additional model to the OOD data.

\subsection{Maximum mean discrepancy for out-of-distribution detection}\label{sec_modelagnostic:MMD}
Another way of approaching out-of-distribution detection from a testing perspective is through a \emph{two-sample test}. Denoting $p_\textup{data}$ the true training data distribution, the goal is to test $\mathcal{H}_0: \tilde{x}_1,\ldots,\tilde{x}_n \sim p_\textup{data}$ against $\mathcal{H}: \tilde{x}_1,\ldots,\tilde{x}_n \not\sim p_\textup{data}$, where the alternative hypothesis $\mathcal{H}$ again is that the test points are OOD.

A popular way of building statistics for two-sample tests is to use a measure of distance between $p_\textup{data}$ and the distribution of $\tilde{x}_1,\ldots,\tilde{x}_n$. The key idea here will be to use the trained generative model to build this measure of distance. To this end, we will use the \emph{maximum mean discrepancy (MMD)} of \textcite{gretton_kernel_2012}, which is a kernel-based measure of distance. Then, $p_\theta$ will be used to specify an appropriate kernel.

More specifically, given a kernel whose feature map is $\Phi : \mathcal{X} \to \mathcal{H}$, the MMD between two distributions $P$ and $Q$ over $\mathcal{X}$ is defined as
\begin{equation}
    \textup{MMD}_\Phi(P,Q) = \| E_{X\sim P}[\Phi(X)] - E_{Y\sim Q}[\Phi(Y)] \|_\mathcal{H}. \hspace{-.1em}
\end{equation}
In our context, the test statistics will be of the form
\begin{equation}
    \label{eq_modelagnostic:mmd}
    \textup{MMD}_\Phi\left(\frac{1}{m}\sum_{i=1}^{m} x_i,\frac{1}{n}\sum_{i=1}^n \tilde{x}_i\right) = \left\|\frac{1}{m}\sum_{i=1}^{m} \Phi(x_i) -  \frac{1}{n}\sum_{i=1}^{n} \Phi(\tilde{x}_i)\right\|_\mathcal{H},
\end{equation}
where $\Phi$ is a kernel feature map built using the generative model and $x_1,\ldots,x_m$ is the training data, i.e.\@ samples from  $p_\textup{data}$. When $\mathcal{H}$ is a simple finite-dimensional Hilbert space and $\Phi$ can be computed easily, then \cref{eq_modelagnostic:mmd} can be computed by going through the data and computing the means in an online fashion.% While the definition seems to imply that we can only perform group-sample OOD detection, it is straightforward to notice that any test statistic of the form of \cref{eq_modelagnostic:mmd} can be used also for single-test OOD detection when $n=1$.%by just substituting  $\frac{1}{n}\sum_{i=1}^{n} \Phi(\tilde{x}_i)$ with simply $\Phi(\tilde{x})$.

@@

\paragraph{The Fisher kernel} An important example of kernel based on a generative model is the \emph{Fisher kernel} of \textcite{jaakkola_exploiting_1999}. The embedding of this kernel is the Fisher score 
\begin{equation}
    \label{eq_modelagnostic:Fisher_score}
    \Phi_\textup{Fisher} (x) = I(\theta)^{-\frac{1}{2}} \nabla \log p_\theta (x),
\end{equation}

and the corresponding reproducing kernel Hilbert space norm is just the $\ell_2$ norm: $|| \cdot || _\mathcal{H} = || \cdot ||_2$. In the case of the Fisher kernel, this means that \cref{eq_modelagnostic:mmd} becomes:
\begin{multline}
    \label{eq_modelagnostic:mmd_fisher}
    \textup{MMD}_{\Phi_\textup{Fisher}}\left(\frac{1}{m} \sum_{i=1}^{m} x_i,\frac{1}{n}\sum_{i=1}^n \tilde{x}_i\right) = \\ \left\|\frac{I(\theta)^{-\frac{1}{2}}}{m} \sum_{i=1}^{m}  \nabla \log p_\theta (x_i) - \frac{I(\theta)^{-\frac{1}{2}}}{n}  \sum_{i=1}^{n}  \nabla \log p_\theta (\tilde{x}_i)\right\|_2.
\end{multline}
We will see later that MMD with a Fisher kernel is closely related to the score statistic. In \cref{appendix_modelagnostic:mahalanobis}, we additionally show that another popular OOD metric known as the \emph{Mahalanobis score} \parencite{lee_simple_2018} can be interpreted as a MMD statistic with a certain Fisher kernel.

\paragraph{The typicality kernel} A very simple approach of embedding the data using $p_\theta$ is to choose $\Phi_\textup{Typical}(x) = \log p_\theta (x)$. Then, MMD is exactly equivalent to the \emph{typicality test statistic} of \textcite{nalisnick_detecting_2019}, although this connection was not explicitly stated  by \textcite{nalisnick_detecting_2019}. Because of this, we call the kernel $k(x,y) = \log p_\theta (x) \cdot \log p_\theta (y)$ the \emph{typicality kernel}.  %We could also call this "new" kernel the "likelihood kernel".
%While $\Phi_\textup{Typical}$ is not as well motivated as the Fisher kernel, we found that it generally gives good results. 
While $\Phi_\textup{Typical}$ is not as well motivated as a kernel as $\Phi_\textup{Fisher}$, the concepts of typicality and typical set can be used to explain unintuitive behaviours of probability distributions in high-dimensional space as highlighted by \textcite{nalisnick_deep_2019}. We also found that using this kernel generally gives good results for OOD tasks. An interesting analysis that we d not consider in this paper would be to study the properties of this kernel.

In general, neither of these two kernels are characteristic, meaning that our MMD can be zero even if the distributions are not identical. This could be solved by combining them with a characteristic kernel, as in \textcite{liu_learning_2020}, at the price of including a new hyperparameter.

@@

%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%  SECTION 3   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Combining different test statistics}
\label{sec_modelagnostic:combination_p_values}
For single-sample OOD detection, \textcite{zhang_understanding_2021} proved that there is not a single statistic that is constantly better compared to all the possible alternatives of interest. For this reason, we believe that using a combination of different test statistics should lead to an overall better OOD detection in settings where a single statistic might fail. Assume we compute $k$ different test statistics $T_1,\dots, T_k$, each testing $\mathcal{H}_0$ against $\mathcal{H}$ as defined in \cref{sec_modelagnostic:parametric,sec_modelagnostic:MMD}. The goal is to combine these different tests into a single statistical test that ideally will perform better than the initial single tests. However, different tests can have different magnitudes and they can differ also in the direction of out-of-distribution detection, i.e.\@ for some statistics having a higher values is associated with being OOD, while for other smaller values are OOD. This makes a combination non-trivial.

\textcite{morningstar_density_2021} proposed the density of states estimator (DoSE) to overcome this problem. They only focused on the single-sample detection task, i.e.\@ $n=1$ following our problem formulation. Their idea is to fit different nonparametric density estimators, such as a kernel-density estimator (KDE) or a one-class support vector machines (SVM), for each different statistic $T_1,\dots, T_k$ by using the values computed on the training set examples. For a single test example, $\tilde{x}_1$, they first compute $T_1,\dots, T_k$ and then combine those statistics by summing the different KDEs log-density. %This approach can be used when combining both one-sided and two-sided test statistic. 
While this approach can be used for any type of statistic, and thus is more general, it uses less prior information. Indeed, if we use only statistics that are truly one-sided, then we assume that a method that leverages the true nature of the statistics should work better. In addition to that, fitting a KDE introduces an additional hyperparameter.

In our work, instead, we propose a different approach and leverage the fact that we use only one-sided test statistics. This setting is a well-studied problem in the literature both for independent \parencite{fisher_statistical_1925, folks_asymptotic_1971} and dependent one-sided test statistics \parencite{brown_400_1975, wilson_harmonic_2019}. All these approaches rely on the computation of $p$-values of each statistic for the test set $\tilde{x}_1,\ldots,\tilde{x}_n$. This corresponds to computing
$p_j = \textup{Pr}(T_j > t_j \mid \mathcal{H}_0)$, i.e.\@ the probability that the $j$'th test is bigger than the observed value under the null hypothesis $\mathcal{H}_0$, where we assume that each $T_j$ has a continuous distribution. Using $p$-values also solves the problem of the statistics having different scales. Indeed, $p$-values transform the different test statistics into the unit interval.

\paragraph{Computation of $p$-values} 
We want to approximate the distribution of the $p$-values $p_1, \ldots, p_k$ of $\tilde{x}_1,\ldots,\tilde{x}_n$ under the null hypothesis $\mathcal{H}_0$. When $\mathcal{H}_0$ is true, then $p_j$ is uniformly distributed on the interval $[0,1]$. To succeed in this, we should be able to compute $p_j = \textup{Pr}(T_j > t_j \mid \mathcal{H}_0)$, therefore we need to estimate the distribution of each statistic $T_j$ under $\mathcal{H}_0$. %As done by \textcite{nalisnick_detecting_2019}, we assume the existence of a validation set $\mathbf{X}'$ that was not used to train our generative model. Instead of bootstrapping $K$ new datasets $\{X^{'}_k\}_{k=1}^{K}$ of size $M$, as they did in their work, we directly evaluate each test statistic $T_j$ on every single validation example. Asymptotically, this is equivalent to creating $K$ new datasets of size $M=1$ when $K \rightarrow \infty$. 
As done by \textcite{nalisnick_detecting_2019}, we assume the existence of a validation set $\mathbf{X}'$ that was not used to train our generative model. From $\mathbf{X}'$ we bootstrap $S$ new datasets $\{\mathbf{X}^{'}_s\}_{s=1}^{S}$ of size $M'$ by using bootstrap resampling. When $n$ is small, for example $n=1$ or $n=2$, where $n=1$ corresponds to single-sample OOD detection, and the validation set is big, a convenient alternative to bootstrapping is to directly evaluate each test statistic $T_j$ on every single validation example. Asymptotically, this is equivalent to creating $S$ new datasets of size $M'=1$ when $S \rightarrow \infty$. In case of $n=2$, i.e.\@ two-samples OOD detection, and a big validation set we can simply bootstrap without resampling.
We then use these values to estimate the empirical distribution function (eCDF) of the considered statistic $T_j$ under $\mathcal{H}_0$. To obtain the $p$-values of test examples $\tilde{x}_1,\ldots, \tilde{x}_n$ for the test statistic $T_j = t_j$, we simply compute $p_j = 1 - \textup{Pr}(T_j < t_j \mid \mathcal{H}_0)$ using the eCDF.
